Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-18 paper/paper.qmd
Methods: severity definitions

@@

IDSA definition of severe CDI based on lab values.
CDC definiton of severe CDI based on disease-related complications [@mcdonald_recommendations_2007].

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitons of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
for all steps of the machine learning analysis.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the Area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

Since the severity labels are imbalanced with different frequencies of severity
for each definition, we calculated balanced precision, the precision expected if
the labels were balanced.
The balanced precision and Area under the balanced precision-recall curve (AUBPRC)
were calculated with Equations 1 and 7 from @wu_improved_2021.

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
A) Decision flow chart to define CDI cases as severe according to the Infectious Diseases Society of America (IDSA) based on lab values,
the occurence of complications due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to CDI with chart review (Attrib).
B) The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition when possible,
and falls back to the All-cause definition when chart review is not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the Area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels available for each definition.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 80% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation performance)
are more important.
Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels available for each definition.
"}
![](figures/feature-importance.png)
:::

>>>>

The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
however, it is easy to collect.

@@

We explore four different ways to define CDI cases as severe or not.
The IDSA definition of severe CDI is based on lab values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
The remaining definitions focus on the occurrence of adverse outcomes, which
may be more clinically relevant.
The all-cause severity definition defines a case as severe if ICU admission,
colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
cause of the adverse event.
The attributable severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
Finally, we defined a pragmatic severity definition that makes use of the
attributable definition when available and falls back to the all-cause definition
when chart review has not been completed, allowing us to use as many samples as
we have available while taking physicians' expert opinions into account where
possible.

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
for all steps of the machine learning analysis.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

Since the severity labels are imbalanced with different frequencies of severity
for each definition, we calculated balanced precision, the precision expected if
the labels were balanced.
The balanced precision and the area under the balanced precision-recall curve
(AUBPRC) were calculated with Equations 1 and 7 from @wu_improved_2021.

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the Infectious
Diseases Society of America (IDSA) based on lab values,
the occurence of complications due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to
CDI with chart review (Attrib).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 80% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
"}
![](figures/feature-importance.png)
:::
